\documentclass[]{article}
\usepackage{caption,subcaption,graphicx,float,url,amsmath,amssymb,tocloft}
\usepackage[hidelinks]{hyperref}
\usepackage[toc,acronym,nonumberlist]{glossaries}
\setacronymstyle{long-short}
\usepackage{glossaries-extra}
\graphicspath{{figs/}} 
\setlength{\cftsubsecindent}{0em}
\setlength{\cftsecnumwidth}{3em}
\setlength{\cftsubsecnumwidth}{3em}
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

%opening
\title{
	\gls{gls:SFI} Origins of Life Course\\
	Peer Review Assignment}
\author{Lawki}

\makeglossaries
\loadglsentries{glossary-entries}

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}

\section{Energetics of metabolism}

\subsection{
	Based on your knowledge of the history of Earth and natural selection, why is methanogenesis more likely to be an early evolutionary metabolism, while aerobic oxidation was a later development?
}

\subsection{
	If aerobic oxidation is more energetically favorable, why would methanogenesis still exist today?}

\section{Life elsewhere}

You have been asked to participate on a mission to locate extraterrestrial life. What geochemical signatures would you want to look for (provide a minimum of 3 with justification) to help you determine if a planet is or has been inhabited in the past, based on the fossil record of the Earth and geochemistries elsewhere in the solar system?

\section{Phylogenic tree building}

\subsection{ Generate a phylogenic tree
	based on a single protein (or
	nucleotide) sequence}

\subsection{ Generate a phylogenic tree
	based on the known taxonomy}

\subsection{ Compare your protein and
	taxonomic trees. Do you notice
	any differences? (Include at
	least 1 difference, or state that
	they are identical)}

\subsection{ What are the challenges with
	building these trees?}

\section{Metabolic Flux}

\subsection{Make a 2D map of the metabolic rate per unit volume as
	a function of cell size on one axis and on the other
	axis.}

\subsection{Simulate the concentration field around a cell}
% end of text 

% glossary
\printglossaries

% bibliography go here

\bibliographystyle{unsrt}
\addcontentsline{toc}{section}{Bibliography}
\bibliography{origins,wikipedia}

\end{document}
